---
title: "UTDEventData: Extracting and aggregating event data from the UTD server"
author: 
- name: "Dr. Patrick T. Brandt"
  affiliation: &utd the University of Texass at Dallas
  email: pbrandt@utdallas.edu
- name: "Dr. Vito D'Orazio"
  affiliation: *utd
  email: dorazio@utdallas.edu
- name: "Hyoungah (Kate) Kim"
  affiliation: *utd
  email: hyoungah.kim@utdallas.edu
- name: "Jared Looper" 
  affiliation: *utd
  email: jrl140030@utdallas.edu
- name: "Micheal J. Shoemate" 
  affiliation: *utd
  email: michael.shoemate@utdallas.edu

package: UTDEventData
date: "`r Sys.Date()`"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{UTDEventData}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r style, echo = FALSE, results = 'asis', eval=FALSE}
BiocStyle::markdown(css.files = c('custom.css'))
```

# Introduction
This R package helps a user to extract a data set from the UTD Event Data server with the country names and time ranges. The project of the UTDEventData R package has not fully completed and is updated everyday. Your comments, feedback, and suggestions are welcome to develop it.
If you have questions in using the package, please contact Kate Kim (<hyoungah.kim@utdallas.edu>) at UTD.  

This package requires you to have an API key to access to the UTD data server. Please find the following link and fill out the form to obtain an API key: <http://eventdata.utdallas.edu/signup>. 

You can install the package from the UTDEVentData github page in R with the following syntax.
```{r, eval = FALSE}
devtools::install_github("KateHyoung/UTDEventData")
```


The UTD Event Data server has the five different event data sets. 

Data Table         | Timeline          |  Further Information
-------------------| :----------------:|--------------------
Phoenix RT         | Oct. 2017 - Today | [OEDA](http://openeventdata.org/)
ICEWS              | 1995 - Sep.2016   | [ICEWS Dataverse](https://dataverse.harvard.edu/dataverse/icews)
Cline Phoenix NYT  | 1945 - 2005       | [Cline Center](http://www.clinecenter.illinois.edu/data/event/phoenix/)
Cline Phoenix FBIS | 1945 - 2005       | [Cline Center](http://www.clinecenter.illinois.edu/data/event/phoenix/)
Cline Phoenix SWB  | 1979 - 2015       | [Cline Center](http://www.clinecenter.illinois.edu/data/event/phoenix/)

Table: (\#tab:dt) Data Table Information 

The web links of each table provides the details of database and their entities and attributes. 

The codes/functions in this package have been tested in the following versions of:

  - Microsoft Open R 3.4.0 
  - R-3.4.3 for Windows (32/64 bit)

# Functions
## Searching functions

There are two main searching functions in this package for users to look up the name of data table and the list of variables each data table contains.  

`DataTables()` returns the all data table names in the UTD server. 
```{r,eval = FALSE}
# Direct way as inputing an API key
DataTables(api_key = " ")
"'PHOENIX_RT', 'CLINE_PHOENIX_SWB', 'CLINE_PHOENIX_FBIS', 'CLINE_PHOENIX_NYT', 'ICEWS'"

# Save an API key as a string value and use it so as not to repeat typing the key string in other functions
k<-"...api key...."
DataTables(k)
"'PHOENIX_RT', 'CLINE_PHOENIX_SWB', 'CLINE_PHOENIX_FBIS', 'CLINE_PHOENIX_NYT', 'ICEWS'"
```

`tableVar()` returns the list of variables in a particulr data table a user specified in the function. For example, the variables in the PHOENIX real-time data can be obtained with the following code. 

```{r, eval = FALSE}
tableVar(api_key="...", table="phoenix_rt")

# with the saved API key to avoid the repeation of API key typing 
k<-"...api key...."
tableVar(k, "phoenix_rt")

tableVar(k, "icews")

tableVar(k, "cline_Phoenix_swb")
```

This function is not case sensitive, so a user may type either lower case or upper case of data table name. However, the full name as returned by `Table()` should be specified in the function. For instance, `"cline_Phoenix_swb"` and `"CLINE_PHOENIX_SWB"` will return the same variables, but `"cline_phenix"` will return noting. 

This function has another feature as enable a user to look up a particular variable. For example, one who may wonder the ICEWS data have a variable named "target" can type the string of the certain variable as follows;

```{r, eval = FALSE }
tableVar(api_key="...", table="icews", lword="target")

k <- "..api key..."
tableVar(k, table="icews", lword="target")
" Target Name"    " Target Sectors", ....
```


## Subsetting function

`pullData()` returns a subset of data from a data table according to the information of country names and time ranges a user provides to the function. Please note the corresponding time period of a certain data in Table \@ref(tab:dt). If the given time range in the function falls outside the timeline of a data table, the function will return `list()` as its result. The format of time range is "YYYYMMDD."     
The country names can be either full names or the ISO-Alpha3 code. Please use a consistent format when you type countries names. We recommend the ISO code format to reduce systemic error in the function.  
The following code is the possible examples you may use to obtain a subset of data. 

```{r, eval = FALSE}
pullData(api_key=" ", table_name="Phoenix_rt", country=list("USA","MEX","SYR","CHN"), start="20171101", end="20171112")

## When you avoid repeating an API key into the function
k<-'api key...'
subset1 <- pullData(k, "phoenix_rt", list('canada','China'), '20171101', '20171102')
subset2 <- pullData(k, "icews", list('can', 'usa'), '20010101','20010110')
subset3 <- pullData(k, 'cline_Phoenix_NYT',list('South Korea','canada'), '19551105','19581215')
```


## Reference class for appying an API key to the functions 

The pacakge has a reference class, named `Table()` for users not to repeat the input of an API key into the searching and subsettng functions. Once a referece class is set, a user does not need to repeatedly put an API key into a function. Some basic usage is as follows;

```{r, eval=FALSE}
# Creating an object
obj<-Table$new()

# Setting an object of an API key
obj$setAPIKey("....")
obj$DataTables()  # returns the available data tables in the UTD server
obj$tableVar("cline_Phoenix_NYT")

# when a user wants to subset real-time data ('phoenix_rt) from 20171101 to 20171102 on MEX(Mexico)
obj$pullData("Phoenix_rt", list("MEX"),start="20171101", end="20171102")
```
You can apply a reference class on the functions of `DataTables()`, `TableVar()`, and `pullData()`. 


# Aggregation functions



# Further information of Event data at UTD

## Spec-real-time server
[Github page for the API access on JetStream at UTD](https://github.com/Sayeedsalam/spec-event-data-server)  
This Github page provides specific information of the direct access to the UTD server managed by Big Data Management and Analytic Lab with Mongo DB query syntax. The query is expressed in JSON format. 



